### 2018-11-09 treestats_paper.tex
covariance notes

\plr{need something different here}
Today's vast quantity of whole-genome sequence
makes it possible to confidently measure quantities
between \emph{individuals} that previously had to be roughly estimated
averaging over populations.
For instance, XXX estimated $F_{ST}$ between XX from allozyme sequence,
but now we know that $F_{ST}$ can be viewed as $1 - t_{within}/t_{total}$ \citep{slatkin_fst}
which can therefore be estimated precisely by $1 - \pi_{\within}/\pi_{total}$.

@@

>>>>

\plr{need something different here}
Today's vast quantity of whole-genome sequence
makes it possible to confidently measure quantities
between \emph{individuals} that previously had to be roughly estimated
averaging over populations.
For instance, XXX estimated $F_{ST}$ between XX from allozyme sequence,
but now we know that $F_{ST}$ can be viewed as $1 - t_{within}/t_{total}$ \citep{slatkin_fst}
which can therefore be estimated precisely by $1 - \pi_{within}/\pi_{total}$.

@@

Let $G$ denote the genotype matrix, so that $G_{ja} \in \{0,1\}$ 
is the genotype of the $a^\text{th}$ individual
at the $j^\text{th}$ site, with `1` denoting the derived state
(we assume here biallelic markers).
Also let $P_j = (1/n) \sum_a G{ja}$ denote the vector of allele frequencies.
Then the \emph{genetic covariance} between samples $a$ and $b$ is
\begin{align*}
    C_{ab} 
        &= \frac{1}{L} \sum_j (G_{ja} - P_j) (G_{jb} - P_j) . %  \\
        % &= \frac{1}{L} \left( (G - P \bone^T)^T (G - P \bone^T) \right)_{ab} .
\end{align*}
This can be computed by letting
$\iw_a = \delta_a$ and $\iw_b = \delta_b$ 
(these weights mark nodes above $a$ and $b$ respectively)
and $\iw_t = \bone / n$
(this weight gives the proportion of samples below each node).
Then, with 
$F(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
we can compute the covariance as
\begin{align*}
    C_{ab} = \frac{1}{2} \bar \site(F, (\iw_a, \iw_b, \iw_t)) .
\end{align*}
This follows, including the factor of two, 
because the summary function applied to the ancestral allele
at a site with derived allele frequencies $p_a$, $p_b$, and $p_t$
in sample $a$, sample $b$, and total, respectively, is
$F(1 - p_a, 1 - p_b, 1 - p_t) = F(p_a, p_b, p_t)$.

Computing the full covariance matrix for a large number of samples may be infeasible.
However, we can efficiently find $y^T C z$ for arbitrary vectors $y$ and $z$.
This works because if we set the initial weights to $y$,
then each allele's weight is equal to the sum of the entries of $y$
corresponding to samples carrying that allele.
Then, if we define 
$F_{yz}(u, v, t) = (u - (\sum_a y_a) t) (v - (\sum_b z_b) t)$,
then
\begin{align} \label{eqn:yCz}
    \frac{1}{2} \bar \site(F_{yz}, (y, z, \iw_t))
        &= \frac{1}{L} \sum_j (\sum_a y_a G_{ja} - (\sum_a y_a) P_j) 
                    (\sum_b z_b G_{jb} - (\sum_b z_b) P_j) \\
        &= y^T C z .
\end{align}

\plr{Hmph: to actually compute $Cz$ we need one weight vector for each sample, again.
    TODO: look up if you can do Krylov just by computing $y^t C z$ for a reasonable number of vectors.  
    We could compute $Cz$ quickly if we could propagate down the tree:
    we'd have $(Cz)_a$ equal to the sum of the weights at all nodes \emph{above} $a$. }

Where does the signal from an eigenvector come from?
Let $y$ be an eigenvector of $C$ with eigenvalue $\lambda$, normalized so that $\|y\|^2 = 1$.
Since $Cy = \lambda y$,
we can write $1 = \sum_a y_a^2 = (1/\lambda) y^T C y$.
However, equation \eqref{eqn:yCz} gives $y^T C y$ as a sum over \emph{sites},
and thus distributes the variance in $G$ associated with $y$
across the tree sequence.
(This is true because of the singular value decomposition of $G$.)
